---
title: "Biostat 203B Homework 3"
subtitle: Due Feb 25 @ 11:59PM
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

Display machine information:
```{r}
sessionInfo()
```

Load database libraries and the tidyverse frontend:
```{r}
suppressPackageStartupMessages(library(dbplyr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
library(DBI)
library(bigrquery)
```

## Q1. Compile the ICU cohort in HW2 from the Google BigQuery database 

Below is an outline of steps.

1. Load the service account token.
```{r}
# path to the service account token 
if (Sys.info()[["nodename"]] == "biostat-203b-teaching-server") {
  # on teaching server
  satoken <- "/mnt/mimiciv/1.0/biostat-203b-2022winter-3fdc2392ac39.json"
} else {
  # on my own Mac laptop
  satoken <- "/Users/huazhou/Documents/github.com/ucla-biostat-203b/2022winter/hw/hw3/biostat-203b-2022winter-3fdc2392ac39.json"
}
# BigQuery authentication using service account
bq_auth(
  path = satoken,
  # email = "mimiciv-bigquery@biostat-203b-2022winter.iam.gserviceaccount.com",
  # scopes = c("https://www.googleapis.com/auth/bigquery",
  #   "https://www.googleapis.com/auth/cloud-platform")
)
```

1. Connect to BigQuery database `mimic4_v1_0_203b` in GCP (Google Cloud Platform), using the billing account `biostat-203b-2022winter`.
```{r}
# Connect to the BigQuery database `biostat-203b-2022winter.mimic4_v1_0_203b`
con <- dbConnect(
    bigrquery::bigquery(),
    project = "biostat-203b-2022winter",
    dataset = "mimic4_v1_0_203b",
    billing = "biostat-203b-2022winter"
)
con
```

2. List all tables in the `mimic4_v1_0_203b` database.
```{r}
dbListTables(con)
```

3. Connect to the `icustays` table.
```{r}
# full ICU stays table
icustays_tble <- tbl(con, "icustays") %>%
  show_query() %>%
  print(width = Inf)
```
We only keep the first ICU stay. Following code is kind of a hack, using the `summarise_all(min)` function. It seems that `slice_min()`, `slice_head()`, `distinct(, .keep_all = TRUE)` don't work with `dbplyr`+`bigrquery` at the moment.
```{r}
icustays_tble <- icustays_tble %>% 
  select(subject_id, intime) %>% 
  group_by(subject_id) %>% 
  summarise_all(min) %>%
  left_join(icustays_tble, by = c("subject_id", "intime")) %>%
  show_query() %>%
  print(width = Inf)
```


4. Connect to the `admissions` table and only keep the patients who have a match in `icustays_tble` (according to `subject_id` and `hadm_id`).
**Answer:**
```{r}
# # TODO
admissions_tble <- tbl(con, "admissions") %>% 
  semi_join(icustays_tble, by = c("subject_id", "hadm_id")) %>% 
  print(width = Inf)
```

5. Connect to the `patients` table and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).
```{r}
# # TODO
patients_tble <- tbl(con, "patients") %>% 
  semi_join(icustays_tble, by = "subject_id")
# collect() # enforce to do the work (query all data (not lazy, so we will see the row count in the printout); put it after summarise ...)
# dbplyr using SQLite is much faster - lazy, not doing work
```

6. Connect to the `labevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the lab items listed in HW2. Only keep the first lab measurements during ICU stay and pivot lab items to become variables/columns.
```{r}
# # TODO
labevents_tble <- tbl(con, "labevents") %>% 
  semi_join(icustays_tble, by = "subject_id") %>% 
  filter(itemid %in% c(50912, 50971, 50983, 50902, 50882, 51221, 51301, 50931, 
                       50960, 50893)) %>% 
  left_join(select(icustays_tble, subject_id, intime), by = "subject_id") %>%
  group_by(subject_id) %>% 
  filter(charttime >= intime) %>% 
  slice_min(charttime) %>% 
  select(subject_id, itemid, valuenum) %>% 
  collect()

labevents_tble$itemid <- recode_factor(labevents_tble$itemid, 
                                       `50912` = "creatinine", 
                                       `50971` = "potassium", 
                                       `50983` = "sodium", 
                                       `50902` = "chloride", 
                                       `50882` = "bicarbonate", 
                                       `51221` = "hematocrit", 
                                       `51301` = "wbcc", 
                                       `50931` = "glucose", 
                                       `50960` = "magnesium", 
                                       `50893` = "calcium")

labevents_tble <- labevents_tble %>% 
  group_by(subject_id, itemid) %>%
  dplyr::mutate(row = row_number()) %>%
  pivot_wider(names_from = "itemid", values_from = "valuenum") %>% 
  select(-row)
```



7. Connect to `chartevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the chart events listed in HW2. Only keep the first chart events during ICU stay and pivot chart events to become variables/columns.
```{r}
# # TODO # keep everything lazy
chartevents_tble <- tbl(con, "chartevents") %>% 
  semi_join(icustays_tble, by = "subject_id") %>% 
  filter(itemid %in% c(220045, 220181, 220179, 223761, 220210)) %>%
  left_join(select(icustays_tble, subject_id, intime), by = "subject_id") %>%
  group_by(subject_id) %>% 
  filter(charttime >= intime) %>% 
  slice_min(charttime) %>% 
  select(subject_id, itemid, valuenum) %>% 
  collect()
  
chartevents_tble$itemid <- recode_factor(chartevents_tble$itemid, 
                                         `220045` = "heart_rate",
                                         `220181` = "mnibp",
                                         `220179` = "snibp",
                                         `223761` = "body_temp",
                                         `220210` = "resp")

chartevents_tble <- chartevents_tble %>% 
  group_by(subject_id, itemid) %>%
  dplyr::mutate(row = row_number()) %>%
  pivot_wider(names_from = "itemid", values_from = "valuenum") %>% 
  select(-row)
```


8. Put things together. This step is similar to Q7 of HW2. Using one chain of pipes `%>%` to perform following data wrangling steps: (i) start with the `icustays_tble` for the first ICU stay of each unique patient, (ii) merge in admissions and patients tables, (iii) keep adults only (age at admission >= 18), (iv) merge in the labevents and chartevents tables, (v) create an indicator for 30-day mortality, (vi) save the final tibble to an `icu_cohort.rds` R data file in the `mimiciv_shiny` folder.
#### do the `collect()` to collect result into `icu_cohort.rds`
#### use `dod` as the time of death
```{r}
# make a directory mimiciv_shiny
if (!dir.exists("mimiciv_shiny")) {
  dir.create("mimiciv_shiny")
}
```

```{r}
# # TODO
# icu_cohort <- icustays_tble %>% ...
icu_cohort <- icustays_tble %>% 
  # merge all dataset by `subject_id`
  left_join(patients_tble, by = "subject_id") %>% 
  left_join(., admissions_tble, by = "subject_id") %>% 
  collect() %>% 
  
  mutate(
    # dod - admittime: time difference in mins
    diff = dod - admittime,
    # thirty_day_mort = 0 if diff > 30 days | is.na(diff)
    thirty_day_mort = ifelse(is.na(diff), 0, ifelse(diff < 30*24*60, 1, 0))
    ) %>% 
  
  # filter the age at admission
  mutate(age_hadm = anchor_age + year(admittime) - anchor_year) %>% 
  filter(age_hadm > 18) %>% 
  select(-diff) %>% 
  
  left_join(., labevents_tble, by = "subject_id") %>% 
  left_join(., chartevents_tble, by = "subject_id")
readr::write_rds(icu_cohort, "icu_cohort.rds")

```

Close database connection and clear workspace.
```{r, eval=F}
dbDisconnect(con)
rm(list = ls())
```

## Q2. Shiny app

Develop a Shiny app for exploring the ICU cohort data created in Q1. The app should reside in the `mimiciv_shiny` folder. The app should provide easy access to the graphical and numerical summaries of variables (demographics, lab measurements, vitals) in the ICU cohort.
```{r, echo = FALSE}
library(ggplot2)
library(shiny)
```

##### Shiny text
```{r, echo = F}
browseURL('https://shiny.rstudio.com/articles/basics.html')
```

# Summary of demographics and lab/vital measurements
```{r, echo = FALSE}
# Define UI for dataset viewer app ----
ui <- fluidPage(

  # App title ----
  titlePanel("Summary"),

  # Sidebar layout with input and output definitions ----
  sidebarLayout(

    # Sidebar panel for inputs ----
    sidebarPanel(

      # Input: Select a variable ----
      selectInput("variable", "Choose a variable:",
                  choices = c("subject_id", "intime", "hadm_id_x", "stay_id",
                              "first_careunit", 
"last_careunit", "outtime", "los", "gender", "anchor_age", "anchor_year", 
"anchor_year_group", "dod", "hadm_id_y", "admittime", "dischtime", 
"deathtime", "admission_type", "admission_location", "discharge_location", 
"insurance", "language", "marital_status", "ethnicity", "edregtime", 
"edouttime", "hospital_expire_flag", "thirty_day_mort", "age_hadm", 
"sodium", "hematocrit", "wbcc", "chloride", "bicarbonate", "glucose", 
"calcium", "potassium", "creatinine", "magnesium", "snibp", "body_temp", 
"heart_rate", "resp", "mnibp")),

      # Input: Specify the number of observations to view ----
      numericInput("obs", "Number of observations to view:", 10),

      # Include clarifying text ----
      helpText("Note: while the data view will show only the specified",
               "number of observations, the summary will still be based",
               "on the full dataset."),

      # Input: actionButton() to defer the rendering of output ----
      # until the user explicitly clicks the button (rather than
      # doing it immediately when inputs change). This is useful if
      # the computations required to render output are inordinately
      # time-consuming.
      actionButton("update", "Update View")

    ),

    # Main panel for displaying outputs ----
    mainPanel(

      # Output: Header + summary of distribution ----
      h4("Summary"),
      verbatimTextOutput("summary"),

      # Output: Header + table of distribution ----
      h4("Observations"),
      tableOutput("view")
    )

  )
)

# Define server logic to summarize and view selected dataset ----
server <- function(input, output) {

  # Return the requested dataset ----
  # Note that we use eventReactive() here, which depends on
  # input$update (the action button), so that the output is only
  # updated when the user clicks the button
  datasetInput <- eventReactive(input$update, {
    switch(input$dataset,
           "rock" = rock,
           "pressure" = pressure,
           "cars" = cars)
  }, ignoreNULL = FALSE)

  # Generate a summary of the dataset ----
  output$summary <- renderPrint({
    dataset <- datasetInput()
    summary(dataset)
  })

  # Show the first "n" observations ----
  # The use of isolate() is necessary because we don't want the table
  # to update whenever input$obs changes (only when the user clicks
  # the action button)
  output$view <- renderTable({
    head(datasetInput(), n = isolate(input$obs))
  })

}

# Create Shiny app ----
shinyApp(ui, server)
```

